---
title: How good is the model?
author: Arthur Andrews
date: 10/1/2024
format: html
editor: source
embed-resources: true
fig-format: svg
fig-width: 10
fig-height: 6
toc: true
toc-location: left
grid: 
  body-width: "9in"
execute:
  echo: false
---

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(lme4)
library(yardstick)
library(splines)
library(reactable)
library(htmltools)
```

```{r}
theme_set(theme_bw())
```

```{r}
load("../../data/ops_models.RData")
```

# Evaluating the player aging model

Previously, we fit a mixed effects model for player ops including

 - splines (to capture the ascent, peak, and descent)
 - random effects (to capture the variety in player performance)

Let's call it the "PAMS model" for "player aging mixed model with splines".

Let's evaluate its accuracy and benchmark against some similar, non-statistical approaches to forecasting player OPS.

## `lme4` model summary

```{r}
form <- formula(model2)

model <- lmer(form, data = hit |> filter(season < 2024))
```

```{r}
model
```


# Test data

We will fit the model on data from 2005 - 2023 and test it it on 2024.

```{r}
#| fig-width: 8
#| fig-height: 2
hit |> 
  count(season) |> 
  mutate(set = if_else(season == 2024, "test", "train") |> factor(c("train", "test"))) |> 
  ggplot() + 
  aes(season, 0, color = set) +
  geom_point(size = 3) +
  theme(
    axis.text.y = element_blank(), 
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    legend.position = "top"
  ) +
  scale_y_continuous(breaks = NULL, minor_breaks = NULL)
```


# Benchmarks

We will compare the model to three simple benchmarks.  These are described in the table.  Hopefully, the statistical model can show a higher quality of fit than the benchmarks.

```{r}
hit <- hit |> 
  group_by(name) |> 
  mutate(
    lagged_ops = lag(ops),
    mean_ops = mean(ops[season < 2024])
  ) |> 
  ungroup() |> 
  mutate(population_ops = mean(ops[season < 2024]))
```

```{r}
hit$pred_ops <- predict(model, hit)
```

```{r}
key <- tribble(
  ~name, ~better_name, ~description,
  "pred_ops", "PAMS model", "mixed model with splines",
  "mean_ops", "player mean ops", "player's mean OPS pre-2024", 
  "lagged_ops", "lagged ops", "player's OPS in 2023",
  "population_ops", "population mean", "mean of all player's OPS"
)
```

# Results



## Quality of fit for the statistical model and three benchmarks

```{r}
hit |> 
  filter(season == 2024) |> 
  select(player = name, contains("ops")) |> 
  pivot_longer(-c(player, ops)) |> 
  group_by(name) |> 
  summarize(
    rmse = rmse_vec(ops, value),
    r2 = rsq_trad_vec(ops, value)
  ) |> 
  arrange(rmse) |>
  left_join(key, by = "name") |> 
  select(name = better_name, description, rmse, r2) |> 
  reactable(
    highlight = TRUE,
    style = css(fontSize = "90%"),
    columns = list(
      rmse = colDef("RMSE", minWidth = 60, format = colFormat(digits = 4)),
      r2 = colDef("R²", minWidth = 60, format = colFormat(digits = 2))
    )
  )
```

